---
title: "How to generate a Minimal Working Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to generate a Minimal Working Example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Please report any issue

We say (one more quote attributed to Albert E.) that a good question solve 80\% of the problem. But Don't wait to solve 80\% of your problem to ask your question.

The only think that can help is to create a Minimal Working Example (MWE). And we provide here a set of MWE you can modify to obtain yours.

Feel free to ask question here: [New Issue for morse R package](https://github.com/pveber/morse/issues)

Here I report a set of minimal example were **each chunk** (piece of code in grey area) work for on its own.

# Loading the R morse library and JAGS and C++

Use the two classical `R` command:

```{r setupMorse, eval=FALSE}
# install.packages(morse) # install the package
library(morse) # load the package
```

Recently, a lot of warning messages have appeared when downloading the `morse` package due to 
an update in its dependency. We try to fix this warnings as soon as possible.

Again, feel free to report such message if such it disturbs you data processing: [New Issue for morse R package](https://github.com/pveber/morse/issues)


The `morse` package is linked to [JAGS](http://mcmc-jags.sourceforge.net/) and C++. These dependencies 

## JAGS

JAGS is the Bayesian sampler used to make inference in survival models.

1. Download JAGS: https://sourceforge.net/projects/mcmc-jags/

[morseDownload.png]!("img/morseDownload.png")

2. Test and Find JAGS

Once installed, JAGS can be lost in the PATH. To help solving this issue, you can use
the excellent package `runjags` which is not within `morse` so you have to isntall it.

```{r testJAGS, eval=FALSE}
install.packages("runjags")
library("runjags")
testjags()
```

We assume hereafter that the morse package is always installed.


The output should look like:

```
You are using R version 4.0.2 (2020-06-22) on a windows machine, with the RStudio GUI
JAGS version 4.3.0 found successfully using the command 'C:/Program Files/JAGS/JAGS-4.3.0/x64/bin/jags-terminal.exe'
The rjags package is installed
```

Otherwise, you can tell your system where is JAGS executable (somewhere in `'C:/Program Files/JAGS/JAGS-4.3.0/x64/bin/jags-terminal.exe'` on windows machines):

```{r pathJAGS, eval=FALSE}
testjags(jags=runjags.getOption('jagspath')) # replace jagspath by the PATH to JAGS
# FOR INSTANCE, on the current machine I do:
testjags(jags=runjags.getOption('C:/Program Files/JAGS/JAGS-4.3.0/x64/bin/jags-terminal.exe'))
```

## C++

C++ is used for running simulations. In R, you should have issue with C++ requirement since it is very well integrated (many R functions are interface to C++ functions).

For now, we do not have any trouble reported on this use. But feel free to report if there is one: [New Issue for morse R package](https://github.com/pveber/morse/issues)

# Survival analysis

## Fit

```{r fitSurv, eval=FALSE}
library(morse)
data("propiconazole")
survData_PRZ = survData(propiconazole)
fit_cstSD = survFit(survData_PRZ, model_type = "SD")
```

### ... with plot

```{r plotSurv, eval=FALSE}
library(morse)
data("propiconazole")
survData_PRZ = survData(propiconazole)
fit_cstSD = survFit(survData_PRZ, model_type = "SD")
plot(fit_cstSD)
```

## LCx

```{r LCxSurv, eval=FALSE}
library(morse)
data("propiconazole")
survData_PRZ = survData(propiconazole)
fit_cstSD = survFit(survData_PRZ, model_type = "SD")
LCx(fit_cstSD, X = 50)
```

### ... with plot

```{r plotLCxSurv, eval=FALSE}
library(morse)
data("propiconazole")
survData_PRZ = survData(propiconazole)
fit_cstSD = survFit(survData_PRZ, model_type = "SD")
LCX_cstSD = LCx(fit_cstSD, X = 50)
plot(LCX_cstSD)
```

### MFX

```{r MFx, eval=FALSE}
library(morse)
data("propiconazole")
survData_PRZ = survData(propiconazole)
fit_cstSD = survFit(survData_PRZ, model_type = "SD")
data_4MFx <- data.frame(time = 1:10,
                        conc = c(0,0.5,8,3,0,0,0.5,8,3.5,0))
MFx_PRZ_cstSD <- MFx(object = fit_cstSD, data_predict = data_4MFx, ode = TRUE)
```

### ... with plot

```{r plotMFx, eval=FALSE}
library(morse)
data("propiconazole")
survData_PRZ = survData(propiconazole)
fit_cstSD = survFit(survData_PRZ, model_type = "SD")
data_4MFx <- data.frame(time = 1:10,
                        conc = c(0,0.5,8,3,0,0,0.5,8,3.5,0))
MFx_PRZ_cstSD <- MFx(object = fit_cstSD, data_predict = data_4MFx, ode = TRUE)
plot(MFx_PRZ_cstSD)
```

### Predict survival probability

```{r predictSurv, eval=FALSE}
library(morse)
data("propiconazole")
survData_PRZ = survData(propiconazole)
fit_cstSD = survFit(survData_PRZ, model_type = "SD")
data_example <- data.frame(
   time = c(1,1.9,2,15,15.1,20),
   conc = c(0,0,20,20,0,0),
   replicate = rep("example", 6)
)
predict_example_NULL = predict_ode(
  object = fit_cstSD,
  data_predict = data_example,
  mcmc_size = 10,
  interpolate_length = NULL)
```

### ... with plot

with few exposure profile time points:

```{r plot1predictSurv, eval=FALSE}
plot(predict_example_NULL)
```

see effect of option `interpolate_length`

```{r plot2predictSurv, eval=FALSE}
predict_example_100 = predict_ode(
  object = fit_cstSD,
  data_predict = data_example,
  mcmc_size = 10,
  interpolate_length = 100)
plot(predict_example_100)
```

and with a very details exposure profile

```{r plotpredictSurv_FOCUS, eval=FALSE}
data("FOCUSprofile")
predict_FOCUS = predict_ode(
  object = fit_cstSD,
  data_predict = FOCUSprofile,
  mcmc_size = 10,
  interpolate_length = NULL)
plot(predict_FOCUS)
```

### Predict number of survival


