---
title: "Library tutorial"
author: "Philippe Ruiz, Philippe Veber, Marie-Laure Delignette-Muller,
Sandrine Charles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
self_contained: no
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  
---

```{r,include=FALSE, echo=FALSE}

knitr::opts_chunk$set(fig.width = 5, 
                      fig.height = 4, 
                      cache = TRUE)

```


```{r, echo=FALSE, cache=TRUE, results='hide'}
library(morsedev)
```

# Introduction

# Species Sensitivity Distribution - SSD

# Reproduction data analyses
  
Before analyse bioassay reproduction data some stages are required. The following example shows all these stages to pass from a raw bioassay dataset to the estimate values of $EC_x$ and parameters. The example is based on the `cadmium1` dataset,
a reproduction and survival dataset of chronic laboratory bioassays with
\emph{Daphnia magna} freshwater invertebrate exposed to five concentrations of
one metal contaminant (cadmium) during 21 days.
  
* Step one, check the structure and the dataset integrity
  
```{r, cache=TRUE}
data(cadmium1)
reproDataCheck(cadmium1, diagnosis.plot = TRUE)
```

The function `reproDataCheck()` performs several tests on the integrity of the
dataset and returns explicit error messages (if necessary). If no error was
detected in the dataset an object of class `reproData` can be created.
If the `reproDataCheck()` function finds an increase of alive individuals with time and if the `diagnosis.plot` argument is `TRUE`, the function plots the number of survivor in function of time per concentration and replicate.
  
* Step two, create a `reproData` object
  
```{r, cache=TRUE}
dat1 <- reproData(cadmium1)
```

The new object contains all the required informations to execute all the reproduction analyses: graphical analyses of data, analyses of reproduction data at all time points.
    
* Step three, plot the `reproData` object
  
`plot()` plots the cmulated number of offsprings depending on the time for all concentrations and replicates.

```{r, cache=TRUE}
plot(dat1)
```

`plot(, target.time)` plots the cumulated number of offsprings depending on the concentration for a fixed time point.
  
```{r, cache=TRUE}
plot(dat1, target.time = 21, addlegend = TRUE)
```

`plot(, concentration)` plots the cumulated number of offsprings depending on the time for a fixed concentration.

```{r, cache=TRUE}
plot(dat1, concentration = 4.36, addlegend = TRUE)
```

and `plot(, conctration, target.time)` compares the cumulated number of offsprings of each replicate for a fixed concentration and a fixed time to the control at the same time.

```{r, cache=TRUE}
plot(dat1, concentration = 4.36, target.time = 21)
```

The `summary()` provides informations on the experimental design.

```{r, cache=TRUE}
dat1
summary(dat1)
```

* Step four, fit an exposure-response model to the reproduction data at target time

Now the dataset with reproduction data at target time is created and it is possible
to estimate the parameters of the log-logistic exposure-response model with the `reproFitTT()` function.
This function calls the `JAGS` program to estimate the model parameters using the bayesian inference.

By choosing the `bestfit` option for the `stoc.part` argument we let the function
choose the best stochastic part of the model.

```{r, results="hide", cache=TRUE}
out <- reproFitTT(dat1, stoc.part = "bestfit",
                  ecx = c(10, 20, 30, 40, 50))
```

The median and 2.5 \% and 97.5 \% of priors and posteriors on estimated parameter distributions and on $EC_{x}$ estimates are given by the `summary()`. The function `plot()` draws the exposure-response curve with the 95 \% credible limits superimposed to experimental data at target-time.

```{r, cache=TRUE}
summary(out)
plot(out, log.scale = TRUE, ci = TRUE)
```

* Step five, check the obseved versus predicted values
  
The function `ppc()` plots the observed versus predicted values with the 95 \% 
credible intervale for each predicted values.

```{r, cache=TRUE}
ppc(out)
```

# Survival data analyses

In this example we fit an exposure-response curve on survival data at final
time.

The two first parts are the sames as the reproduction analysis.

* Step one, check the structure and the dataset integrity
   
```{r, cache=TRUE}
data(cadmium1)
survDataCheck(cadmium1, diagnosis.plot = TRUE)
```

* Step two, create a `survData` object
  
```{r, cache=TRUE}
dat2 <- survData(cadmium1)
dat2
```

* Step three, plot the `survData` object
  
`plot()` plots the number of surviving individuals depending on the time for all
concentrations and replicates.

```{r, cache=TRUE}
plot(dat2)
```

`plot(, target.time)` plots the number of surviving individuals depending on the
concentration for a fixed time point.
  
```{r, cache=TRUE}
plot(dat2, target.time = 21, addlegend = TRUE)
```

`plot(, concentration)` plots the number of surviving individuals depending on the time for a fixed concentration.

```{r, cache=TRUE}
plot(dat2, concentration = 4.36, addlegend = TRUE)
```

and `plot(, conctration, target.time)` compares the number of surviving individuals of each replicate for a fixed concentration and a fixed time to the control at the same time.

```{r, cache=TRUE}
plot(dat2, concentration = 4.36, target.time = 21)
```

The `summary()` provides informations on the experimental design.

```{r, cache=TRUE}
dat2
summary(dat2)
```


* Step four, fit an exposure-response model to the survival data at final time

Now the dataset with survival data at target time is created and it is possible
to estimate the parameters of the log-logistic exposure-response model with the `survFitTT()` function.
This function calls the `JAGS` program to estimate the model parameters using the bayesian inference.

```{r, results="hide", cache=TRUE}
out2 <- survFitTT(dat2,
                  lcx = c(10, 20, 30, 40, 50))
```

The median and 2.5 \% and 97.5 \% of priors and posteriors on estimated parameter distributions and on $LC_{x}$ estimates are given by the `summary()`. The function `plot()` draws the exposure-response curve with the 95 \% confidence interval on observed data.

```{r, cache=TRUE}
summary(out2)
plot(out2, log.scale = TRUE, ci = TRUE)
```
  
* Step five, check the obseved versus predicted values
  
```{r,cache=TRUE}
ppc(out2)
```
