---
title: "morse tutorial"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
self_contained: no
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}

---

```{r,include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.width = 7,
                      fig.height = 4,
                      cache = TRUE)
```


```{r, echo=FALSE, results='hide'}
library(morse)
library(dplyr)
```

The package `morse` is devoted to the analysis of data from standard toxicity
tests. It provides a simple workflow to explore/visualize a dataset, and
compute estimations of risk assessment indicators. This document illustrates
a typical use of `morse` on survival and reproduction data, which can be followed
step-by-step to analyze new datasets.

# Survival data analysis at target time (TT)

The following example shows all the steps to perform survival analysis on
standard toxicity test data and to produce estimated values of the $LC_x$.
We will use a dataset of the library named `cadmium2`, which contains both
survival and reproduction data from a chronic laboratory toxicity test. In this
experiment, snails were exposed to six concentrations of a metal contaminant
(cadmium) during 56 days.

## Step 1: check the structure and the dataset integrity

The data from a survival toxicity test should be gathered in a `data.frame` with a
specific layout. This is documented in the paragraph on `survData` in the
reference manual, and you can also inspect one of the datasets provided in
the package (e.g., `cadmium2`). First, we load the dataset and use the function
`survDataCheck` to check that it has the expected layout:

```{r step1TT, cache=TRUE}
data(cadmium2)
survDataCheck(cadmium2)
```

The output `## No message` just informs that the dataset is well-formed.

## Step 2: create a `survData` object

The class `survData` corresponds to \emph{validated} survival data and is the
basic layout used for the subsequent operations. Note that if the
call to `survDataCheck` reports no error (i.e., `## No message`), it is guaranteed that `survData`
will not fail.

```{r step2TT, cache=TRUE}
dat <- survData(cadmium2)
head(dat)
```

## Step 3: visualize your dataset

The function `plot` can be used to plot the number of surviving individuals
as a function of time for all concentrations and replicates.

```{r step3TT, cache=TRUE}
plot(dat, pool.replicate = FALSE)
```


Two graphical styles are available, `"generic"` for standard `R` plots or
`"ggplot"` to call package `ggplot2` (default). If argument `pool.replicate`
is `TRUE`, datapoints at a given time-point and a given concentration are pooled
and only the mean number of survivors is plotted. To observe the full
dataset, we set this option to `FALSE`.

By fixing the concentration at a (tested) value, we can visualize one subplot
in particular:

```{r, cache=TRUE}
plot(dat, concentration = 124, addlegend = TRUE,
     pool.replicate = FALSE, style ="generic")
```


We can also plot the survival rate, at a given time-point, as a function of 
concentration, with binomial confidence intervals around the data. This is 
achieved by using function `plotDoseResponse` and by fixing the option 
`target.time` (default is the end of the experiment).

```{r, cache=TRUE}
plotDoseResponse(dat, target.time = 21, addlegend = TRUE)
```

Function `summary` provides some descriptive statistics on the experimental design.

```{r, cache=TRUE}
summary(dat)
```

## Step 4: fit an exposure-response model to the survival data at target time

Now we are ready to fit a probabilistic model to the survival data, in order to
describe the relationship between the concentration in chemical compound and survival rate
at the target time. Our model assumes this latter is a log-logistic function of
the former, from which the package delivers estimates of the parameters.
Once we have estimated the parameters, we can then calculate the $LC_x$ values
for any $x$. All this work is performed by the `survFitTT` function, which requires
a `survData` object as input and the levels of $LC_x$ we want:

```{r step4TT, results="hide", cache=TRUE}
fit <- survFitTT(dat,
                 target.time = 21,
                 lcx = c(10, 20, 30, 40, 50))
```

The returned value is an object of class `survFitTT` providing the estimated
parameters as a posterior[^1] distribution, which quantifies the uncertainty on
their true value. For the parameters of the models, as well as for the $LC_x$
values, we report the median (as the point estimated value) and the 2.5 \% and 97.5 \%
quantiles of the posterior (as a measure of uncertainty, a.k.a. credible
intervals). They can be obtained by using the `summary` method:


```{r step4TTsummary, cache=TRUE}
summary(fit)
```

If the inference went well, it is expected that the difference between
quantiles in the posterior will be reduced compared to the prior, meaning
that the data were helpful to reduce the uncertainty on the true value of
the parameters. This simple check can be performed using the summary function.

The fit can also be plotted:
```{r step4TTplot, cache=TRUE}
plot(fit, log.scale = TRUE, adddata = TRUE,   addlegend = TRUE)
```

This representation shows the estimated relationship between concentration of
chemical compound and survival rate (orange curve). It is computed by choosing for each
parameter the median value of its posterior. To assess the uncertainty on this
estimation, we compute many such curves by sampling the parameters
in the posterior distribution. This gives rise to the grey band, showing for
any given concentration an interval (called credible interval) containing the
survival rate 95% of the time in the posterior distribution. The experimental
data points are represented in black and correspond to the observed survival
rate when
pooling all replicates. The black error bars correspond to a 95% confidence
interval, which is another, more straightforward way to bound the most
probable value of the survival rate for a tested concentration. In favorable
situations, we expect that the credible interval around the estimated curve
and the confidence interval around the experimental data largely overlap.

A similar plot is obtained with the style `"generic"`:
```{r, cache=TRUE}
plot(fit, log.scale = TRUE, style = "generic", adddata = TRUE, addlegend = TRUE)
```


Note that `survFitTT` will warn you if the estimated $LC_{x}$ lie outside the
range of tested concentrations, as in the following example:

```{r wrongTT, results="hide", cache=TRUE}
data("cadmium1")
doubtful_fit <- survFitTT(survData(cadmium1),
                       target.time = 21,
                       lcx = c(10, 20, 30, 40, 50))
plot(doubtful_fit, log.scale = TRUE, style = "ggplot", adddata = TRUE,
     addlegend = TRUE)
```

In this example, the experimental design does not include sufficiently high
concentrations, and we are missing measurements that would have a major
influence on the final estimation. For this reason this result should be
considered unreliable.


## Step 5: validate the model with a posterior predictive check

The fit can be further validated using so-called posterior
predictive checks: the idea is to plot the observed values against the
corresponding estimated predictions, along with their 95% credible
interval. If the fit is correct, we expect to see 95% of the data
inside the intervals.

```{r step5TT,cache=TRUE, results="hide"}
ppc(fit)
```

In this plot, each black dot represents an observation made at a given
concentration, and the corresponding number of survivors at target time is given by
the value on the *x-axis*. Using the concentration and the fitted model, we
can produce the corresponding prediction of the expected number of
survivors at that concentration. This prediction is given by the
*y-axis*. Ideally observations and predictions should coincide, so we'd
expect to see the black dots on the points of coordinate $Y = X$. Our
model provides a tolerable variation around the predited mean value
as an interval where we expect 95% of the dots to be in average. The
intervals are represented in green if they overlap with the line $Y=X$,
and in red otherwise.

# Survival analysis using a toxico-kinetic toxico-dynamic (TK-TD) model

The steps for a TK-TD data analysis are absolutely analogous to what we
described for the analysis at target time. Here the goal is to estimate the relationship
between chemical compound concentration, time and survival rate.

## TK-TD model with constant exposure concentrations

Here is a typical session to analyse concentration-dependent time-course data
using the so-called ["Stochastic Death" (SD) model](modelling.pdf):

```{r TKTDcst, cache=TRUE}
# (1) load dataset
data(propiconazole)

# (2) check structure and integrity of the dataset
survDataCheck(propiconazole)

# (3) create a `survData` object
dat <- survData(propiconazole)

# (4) represent the number of survivors as a function of time
plot(dat, pool.replicate = FALSE)

# (5) check information on the experimental design
summary(dat)
```

### TK-TD model "SD"

To fit the *Stochastic Death* model, we have to specify the `model_type` as `"SD"`:

```{r fitcstSD, cache=TRUE, echo=TRUE}
# (6) fit the TK-TD model SD
fit_cstSD <- survFit(dat, quiet = TRUE, model_type = "SD")
```

Then, the `summary` function provides parameters estimates as medians and 95\% credible intervals.

```{r, cache=TRUE}
# (7) summary of parameters estimates
summary(fit_cstSD)
```

The `plot` function provides a representation of the fitting for each replicates

```{r, cache=TRUE}
plot(fit_cstSD)
```

Original data can be added by using the option `adddata = TRUE`

```{r, cache=TRUE}
plot(fit_cstSD, adddata = TRUE)
```

A posterior predictive check is also possible using function `ppc`:

```{r, cache=TRUE}
ppc(fit_cstSD)
```

Compared to the target time analysis, TK-TD modelling allows to compute and plot the lethal concentration for any *x* percentage and at any time-point. The chosen time-point can be specified with `time_LCx`, by default the maximal time-point in the dataset is used.

```{r cstSDLCx, cache=TRUE}
# LC50 at the maximum time-point:
LCx_cstSD <- LCx(fit_cstSD, X = 50)
plot(LCx_cstSD)

# LC50 at time = 2
LCx(fit_cstSD, X = 50, time_LCx = 2) %>% plot()
## Note the use of the pipe operator, `%>%`, which is a powerful tool for clearly expressing a sequence of multiple operations.
## For more information on pipes, see: http://r4ds.had.co.nz/pipes.html
```

Warning messages are returned when the range of concentrations is not appropriated for one or more LCx calculation(s). 

```{r}
# LC50 at time = 15
LCx(fit_cstSD, X = 50, time_LCx = 15) %>% plot()
```


### TK-TD model "IT"

The *Individual Tolerance* (IT) model is a variant of TK-TD survival
analysis. It can also be used with `morse` as demonstrated hereafter. For the *IT* model, we have to specify the `model_type` as `"IT"`:

```{r fitcstIT, echo=TRUE, cache=TRUE}
fit_cstIT <- survFit(dat, quiet = TRUE, model_type = "IT")
```

We can first get a summary of the estimated parameters:

```{r, cache=TRUE}
summary(fit_cstIT)
# OR
fit_cstIT$estim.par
```

```{r, cache=TRUE}
plot(fit_cstIT, adddata= TRUE)
```

```{r, cache=TRUE}
ppc(fit_cstIT)
```

```{r cstITLCx, cache=TRUE}
# LC50 at the maximum time-point:
LCx_cstIT <- LCx(fit_cstIT, X = 50)
plot(LCx_cstIT)

# LC50 at time = 2
LCx(fit_cstIT, X = 50, time_LCx = 2) %>% plot()

# LC30 at time = 15
LCx(fit_cstIT, X = 30, time_LCx = 15) %>% plot()
```

## TK-TD model under time-variable exposure concentration

Here is a typical session fitting an SD or an IT model for a dataset
under time-variable exposure scenario.

```{r TKTDvar, cache=TRUE}
# (1) load dataset
data("propiconazole_pulse_exposure")

# (2) check structure and integrity of the dataset
survDataCheck(propiconazole_pulse_exposure)

# (3) create a `survData` object
dat <- survData(propiconazole_pulse_exposure)

# (4) represent the number of survivor as a function of time
plot(dat)

# (5) check information on the experimental design
summary(dat)
```

### TK-TD model "SD"

```{r fitvarSD, echo=TRUE, cache=TRUE}
# (6) fit the TK-TD model SD
fit_varSD <- survFit(dat, quiet = TRUE, model_type = "SD")
```

```{r, cache=TRUE}
# (7) summary of the fit object
summary(fit_varSD)
```

```{r, cache=TRUE}
plot(fit_varSD)
```

```{r, cache=TRUE}
ppc(fit_varSD)
```

```{r varSDLCx, cache=TRUE}
# LC50 at time = 4
LCx_varSD <- LCx(fit_varSD, X = 50, time_LCx = 4, conc_range = c(0,100))
plot(LCx_varSD)

# LC50 at time = 30
LCx(fit_varSD, X = 50, time_LCx = 30,  conc_range = c(0,100)) %>% plot()
```

### TK-TD model "IT"

```{r fitvarIT, echo=TRUE, cache=TRUE}
# fit a TK-TD model IT
fit_varIT <- survFit(dat, quiet = TRUE, model_type = "IT")
```

```{r, cache=TRUE}
# (7) summary of the fit object
summary(fit_varIT)
```

```{r, cache=TRUE}
plot(fit_varIT)
```

```{r, cache=TRUE}
ppc(fit_varIT)
```

```{r varITLCx, cache=TRUE}
# LC50 at time = 4
LCx(fit_varIT, X = 50, time_LCx = 4, conc_range = c(0,200)) %>% plot()

# LC50 at time = 30
LCx(fit_varIT, X = 50, time_LCx = 30, conc_range = c(0,100)) %>% plot()
```

## Prediction

GUTS models can be used to simulate the survival of the organisms using any exposure patterns, using the
calibration done with the function `survFit` from observed data.
The function of prediction is called `predict`and returns an object of class `survFitPredict`.


```{r predict, cache=TRUE}
# (1) upload or build a data frame with the exposure profile
# argument `replicate` is used to provide several profiles of exposure
data_4prediction <- data.frame(time = c(1:10, 1:10),
                               conc = c(c(21,19,18,23,20,14,25,8,13,5),
                                        c(0,0,40,0,0,0,40,0,0,0)),
                               replicate = c(rep("random", 10), rep("pulse", 10)))

# (2) Use the fit on constant exposure eropiconazole with model SD (see previously)
predict_PRZ_cstSD_4pred <- predict(object = fit_cstSD, data_predict = data_4prediction)
```

Then, from a `survFitPredict`, there is the function `plot`:

```{r predictPlot, cache=TRUE}
# (3) Plot the predicted survival rate over the new exposure profiles.
plot(predict_PRZ_cstSD_4pred)
```

## Multiplication factors: a 'margin of safety'

Using predictions functions, GUTS models can be used to simulate the survival of the organisms exposed to an observed exposure pattern. Hopefully, in general this real exposure profile does not result in any related mortality, but a critical question is to know how far the exposure profile is from adverse effect, that is a "margin of safety".

This idea is then to multiply the concentration in the real profile by a "multiplication factor", denoted $MF_x$, resulting in $x\%$ (classically $10\%$ or $50\%$) of additional death at a specified time (for instance, at the end of the exposure period ).

The multiplication factor $MF_x$ then informs the "margin of safety", used to assess if the risk is considered acceptable or not.

Computing an $MF_x$ is easy with the function `MFx`. It only requires the object `survFit` and the exposure profile, argument `data_predict` in the function. The chosen percentage of survival reduction is specified with argument `X`, the default is $10$, and the chosen time-point can be specified with `time_MFx`, by default the maximal time-point in the dataset is used. 

There is no explicit formulation of $MF_x$ (at least for the GUTS-SD model), so the `accuracy` argument can be used to change the accuracy of the convergence level.


```{r MFx_compt, cache=TRUE}
# (1) upload or build a data frame with the exposure profile
data_4MFx <- data.frame(time = 1:10,
                        conc = c(0,0.5,8,3,0,0,0.5,8,3.5,0))

# (2) Use the fit on constant exposure eropiconazole with model SD (see previously)
MFx_PRZ_cstSD_4MFx <- MFx(object = fit_cstSD, data_predict = data_4MFx)
# As the computing can be long, the function print the `accuracy` for each step of tested multiplication factor.
```

Then, we can plot the survival rate with both initial profile, and the initial profile multiplied by the obtained multiplicationf factor.

```{r MFx_plot, cache=TRUE}
# (3) Plot the survival rate with the multiplication factor.
plot(MFx_PRZ_cstSD_4MFx)
```

In this specific case the x-axis need to be log-scaled, what is possible by setting the option `log_scale = TRUE`:

```{r MFx_plotLog, cache=TRUE}
plot(MFx_PRZ_cstSD_4MFx, log_scale = TRUE)
```

To compare the initial survival rate (i.e., a multiplication factor set to 1), and the optimal multiplication factor to obtain a reduction of $x\%$ of survival (provided with argument `X`), we can use the option `x_variable =  "Time"`

```{r MFx_plotTime, cache=TRUE}
# option 'x_variable' allow to vizualize difference of survival with and without multiplication factor.
plot(MFx_PRZ_cstSD_4MFx, x_variable =  "Time")
```

Here is an example with a 10 percent Multiplication Factor:

```{r MFx_x10, cache=TRUE}
MFx_PRZ_cstSD_4MFx_x10 <- MFx(object = fit_cstSD, data_predict = data_4MFx, X = 10)
plot(MFx_PRZ_cstSD_4MFx_x10, log_scale = TRUE)
```

Multiplication factor is also available for the GUTS IT model (option `quiet = TRUE` remove the output):

```{r MFx_IT, cache=TRUE}
# (2) Use the fit on constant exposure propiconazole with model IT. No computing of 
MFx_PRZ_cstIT_4pred <- MFx(object = fit_cstIT, data_predict = data_4MFx, time_MFx = 4, quiet = TRUE)

# (3) Plot the survival rate with the multiplication factor.
plot(MFx_PRZ_cstIT_4pred, log_scale = TRUE)
```


```{r MFx_ITplot, cache=TRUE}
MFx_PRZ_cstIT_4pred <- MFx(object = fit_cstIT, X=10, hb_value = FALSE, data_predict = data_4MFx, time_MFx = 4)

plot(MFx_PRZ_cstIT_4pred, log_scale = TRUE)
```


# Reproduction data analysis at target-time

The steps for reproduction data analysis are absolutely analogous to what we
described for survival data. Here, the aim is to estimate the relationship
between the chemical compound concentration and the reproduction rate per individual-day.

Here is a typical session:
```{r, cache=TRUE}
# (1) load dataset
data(cadmium2)

# (2) check structure and integrity of the dataset
reproDataCheck(cadmium2)

# (3) create a `reproData` object
dat <- reproData(cadmium2)

# (4) represent the cumulated number of offspring as a function of time
plot(dat, concentration = 124, addlegend = TRUE, pool.replicate = FALSE)

# (5) represent the reproduction rate as a function of concentration
plotDoseResponse(dat, target.time = 28)

# (6) check information on the experimental design
summary(dat)

# (7) fit a concentration-effect model at target-time
fit <- reproFitTT(dat, stoc.part = "bestfit",
                  target.time = 21,
                  ecx = c(10, 20, 30, 40, 50),
                  quiet = TRUE)
summary(fit)
```

```{r, cache=TRUE}
plot(fit, log.scale = TRUE, adddata = TRUE,
     cicol = "orange",
     addlegend = TRUE)
```

```{r, cache=TRUE}
ppc(fit)
```

As in the survival analysis, we assume that the reproduction rate per individual-day
is a log-logistic function of the concentration. More details and parameter
signification can be found in the [modelling vignette](modelling.pdf).

## Model comparison

For reproduction analyses, we compare one model which neglects the inter-individual
variability (named "Poisson") and another one which takes it into account
(named "gamma Poisson"). You can choose either one or the other with the option `stoc.part`.
Setting this option to `"bestfit"`, you let `reproFitTT` decides which models fits the
data best. The corresponding choice can be seen by calling the `summary` function:

```{r, cache=TRUE}
summary(fit)
```
When the gamma Poisson model is selected, the summary shows an additional
parameter called `omega`, which quantifies the inter-individual variability
(the higher `omega` the higher the variability).


## Reproduction data and survival functions

In `morse`, reproduction datasets are a special case of survival datasets: a
reproduction dataset includes the same information as in a survival dataset plus
the information on reproduction outputs. For that reason, the S3 class `reproData`
inherits from the class `survData`, which means that any operation on a `survData`
object is legal on a `reproData` object. In particular, in order to use the plot function
related to the survival analysis on a `reproData` object, we can use `survData` as a
conversion function first:

```{r, cache=TRUE}
dat <- reproData(cadmium2)
plot(survData(dat))
```


[^1]: In Bayesian inference, the parameters of a model are estimated
from the data starting from a so-called *prior*, which is a probability
distribution representing an initial guess on the true parameters, before
seing the data. The *posterior* distribution represents the uncertainty on
the parameters after seeing the data and combining them with the prior. To obtain
a point estimate of the parameters, it is typical to compute the mean or median
of the posterior. We can quantify the uncertainty by reporting the standard
deviation or an inter-quantile distance from this posterior distribution.
